\sysfull

### 2018-04-17 paper/paging.tex
abstract, and had to move figures to fix layout
expand algorithm 3 and 4 to make them more readable, sill within the pages limit
remove the red text on top

\textcolor{red}{[additional motivation for coala memory virtualization] we should emphasize that Coala programming model removes the burden of identifying the write-after-read dependencies from the programmer  as compared to Chain. And it does not require a specific compiler that cannot consider a runtime dependent behavior like Interrupts service handling to guarantee memory protection, as in Alpaca.}

@@

%\begin{algorithm}
\begin{wrapfigure}{t}{0.5\textwidth}
\fbox{\parbox{\linewidth}{
	\captionof{algorithm}{\texttt{PageFault}(tag $t$)}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \State \LeftComment{$i_\text{victim}$: index to free slot or page to be evicted, initially 0}
        \State \LeftComment{\texttt{\underline{shadowCount}}: non-volatile, 0 on boot unless in phase 2 commit}
        \State $t_\text{victim} \gets \Call{Tag}{\texttt{working}[i_\text{victim}]}$
		\If {\Call{isDirty}{working[$i_\text{victim}$]} }	\Comment{Evicting a dirty page?}
		\State $\texttt{\underline{shadow}}[t_\text{victim}] \xleftarrow{\text{DMA}} \texttt{working}[i_\text{victim}]$
            \Comment{Swap out the page}
        \If {$t_\text{victim} \not\in \texttt{\underline{shadowList}}$} \Comment{Keep track for quick traversal}
            \State $\texttt{\underline{shadowList}}[\texttt{shadowCount++}] \gets t_\text{victim}$
        \EndIf
		\EndIf
		\If {\Call{exists}{\underline{shadow}[$t$]}} \Comment{Page swapped out earlier?}
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
            \Comment{Get from commit buffer}
		\Else
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{store}}[t]$
            \Comment{Get from backing store}
		\EndIf 
        \State $i_\text{victim} \gets (i_\text{victim} + 1) \mod |\texttt{pageBufLength}|$ \Comment{FIFO evict}
%	\end{algorithmic}
%\end{algorithm}
	\end{algorithmic}}}
\end{wrapfigure}

If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a 
\emph{victim page}) preserving updates made to that page, and to copy the
accessed page of protected data into the working buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
a \emph{first-in-first-out} (FIFO) replacement policy (Lines 2 and 12). 
%
While we opted for the simplicity of FIFO, any other replacement policy, such
as Least Recently Used (LRU) could work in its place.
%
If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty.  If the page being swapped out is not dirty and the page
being accessed had not been dirtied and swapped out to the \texttt{shadow}
buffer since the last reboot, then handling a page fault requires copying the
accessed page from its home in the  \texttt{\underline{store}} into the
\texttt{working} buffer (Line 11).  Otherwise, the page fault handler must take
the following extra steps.
%
Before swapping in the accessed page, \sys must first preserve the updated
values in the dirty page so that they can be committed back to their locations
in memory when the task commits (but not earlier).  To preserve the dirty
values, \sys copies the victim page to the non-volatile
\texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a
non-volatile commit list (Lines 6-7). \sys uses the commit list to iterate over shadow pages 
at commit time, to copy the contents of shadow pages back to the main memory {\tt store}. 
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last
power failure (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

%\begin{algorithm}[t]
\begin{wrapfigure}{t}{0.5\textwidth}
\fbox{\parbox{\linewidth}{
	\captionof{algorithm}{Two-phase commit}
	%\caption{Two-phase commit}
	\label{algo:commit}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \Procedure{CommitPhase1}{} \Comment{On last coalesced task}
            \For {$i \in 0..|\texttt{working}|-1$}
                \State $t \gets \Call{Tag}{\texttt{working}[i]}$
                \State $\texttt{\underline{shadow}}[\Call{Tag}{\texttt{working}[i]}] \xleftarrow{\text{DMA}} \texttt{working}[i]$
                \State $\texttt{\underline{shadowList}}[\texttt{\underline{shadowCount}}] \gets t$
                \State $\texttt{\underline{shadowCount}} \gets \texttt{\underline{shadowCount}} + 1$
            \EndFor
            \State \Call{CommitPhase2}{}
        \EndProcedure
        \Procedure{CommitPhase2}{} \Comment{\texttt{\underline{commitIndex}} 0 on first boot}
            \State $\texttt{\underline{committing}} \gets \textbf{true}$
            \While{\texttt{\underline{commitIndex}} < \texttt{\underline{shadowCount}}}
                \State $t \gets \texttt{\underline{shadowList}}[\texttt{\underline{commitIndex}}]$
                \State \Call{commitToStore}{\texttt{\underline{shadow}}[t]}
                \State $\texttt{\underline{commitIndex}} \gets \texttt{\underline{commitIndex}} + 1$
            \EndWhile
            \State $\texttt{\underline{shadowCount}} \gets 0$
            \State $\texttt{\underline{commitIndex}} \gets 0$
            \State $\texttt{\underline{committing}} \gets \textbf{false}$
        \EndProcedure
        \Procedure{OnBoot}{} \Comment{Invoked on every boot}
            \If { \texttt{\underline{committing}} } \Call{CommitPhase2}{}
            \EndIf 
            \State $\texttt{\underline{shadowCount}} \gets 0$, $i_\text{victim} \gets 0$
        \EndProcedure
%	\end{algorithmic}
%\end{algorithm}
	\end{algorithmic}}}
\end{wrapfigure}

@@

\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only
ever executes using consistent protected data. During task execution,
modifications to protected data do not affect their original memory locations,
because a task reads and writes the volatile \texttt{working} buffer only, and
modified pages are kept in \texttt{\underline{shadow}} until commit.  A power
failure erases the contents of the page buffer, preventing a re-executing task
from observing updates in the page buffer from a previous execution attempt.
Clearing \texttt{\underline{shadowCount}} as part of the second phase commit
(Line 14), ensures that all accesses to protected variables in subsequent tasks
correctly access their original memory locations in \texttt{\underline{store}}.
\sys assumes that the write to a non-volatile word in memory is atomic.
%
%The non-volatile commit bit persists
%across power failures and \sys repeatedly attempts to commit until it succeeds,
%despite arbitrary power failures.
\subsection{Dynamic Paging of \sys}

>>>>

% \textcolor{red}{[additional motivation for coala memory virtualization] we should emphasize that Coala programming model removes the burden of identifying the write-after-read dependencies from the programmer  as compared to Chain. And it does not require a specific compiler that cannot consider a runtime dependent behavior like Interrupts service handling to guarantee memory protection, as in Alpaca.}

@@

If the \texttt{working} buffer is full, a page fault on a memory access
requires \sys to swap out one of the pages in the volatile buffer (a 
\emph{victim page}) preserving updates made to that page, and to copy the
accessed page of protected data into the working buffer.
Algorithm~\ref{algo:pagefault} shows how \sys handles a page fault. A victim
page in the \texttt{working} buffer is selected for eviction according to
a \emph{first-in-first-out} (FIFO) replacement policy (Lines 2 and 12). 
%
While we opted for the simplicity of FIFO, any other replacement policy, such
as Least Recently Used (LRU) could work in its place.

% \begin{algorithm}
% \begin{wrapfigure}{t}{0.5\textwidth}
% \fbox{\parbox{}{
% 	\captionof{algorithm}{\texttt{PageFault}(tag $t$)}
% 	\label{algo:pagefault}
% 	\scriptsize
\begin{algorithm}
	\caption{\texttt{PageFault}(tag $t$)}
	\label{algo:pagefault}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \State \LeftComment{$i_\text{victim}$: index to free slot or page to be evicted, initially 0}
        \State \LeftComment{\texttt{\underline{shadowCount}}: non-volatile, 0 on boot unless in phase 2 commit}
        \State $t_\text{victim} \gets \Call{Tag}{\texttt{working}[i_\text{victim}]}$
		\If {\Call{isDirty}{working[$i_\text{victim}$]} }	\Comment{Evicting a dirty page?}
		\State $\texttt{\underline{shadow}}[t_\text{victim}] \xleftarrow{\text{DMA}} \texttt{working}[i_\text{victim}]$
            \Comment{Swap out the page}
        \If {$t_\text{victim} \not\in \texttt{\underline{shadowList}}$} \Comment{Keep track for quick traversal}
            \State $\texttt{\underline{shadowList}}[\texttt{shadowCount++}] \gets t_\text{victim}$
        \EndIf
		\EndIf
		\If {\Call{exists}{\underline{shadow}[$t$]}} \Comment{Page swapped out earlier?}
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{shadow}}[t]$
            \Comment{Get from commit buffer}
		\Else
		\State $\texttt{working}[i_\text{victim}] \xleftarrow{\text{DMA}} \texttt{\underline{store}}[t]$
            \Comment{Get from backing store}
		\EndIf 
        \State $i_\text{victim} \gets (i_\text{victim} + 1) \mod |\texttt{pageBufLength}|$ \Comment{FIFO evict}
	\end{algorithmic}
\end{algorithm}

% 	\end{algorithmic}
% \end{algorithm}
	% \end{algorithmic}}}
% \end{wrapfigure}

If a task modified any byte in the victim page (i.e., using \texttt{WP}), then
the page is dirty.  If the page being swapped out is not dirty and the page
being accessed had not been dirtied and swapped out to the \texttt{shadow}
buffer since the last reboot, then handling a page fault requires copying the
accessed page from its home in the  \texttt{\underline{store}} into the
\texttt{working} buffer (Line 11).  Otherwise, the page fault handler must take
the following extra steps.
%
Before swapping in the accessed page, \sys must first preserve the updated
values in the dirty page so that they can be committed back to their locations
in memory when the task commits (but not earlier).  To preserve the dirty
values, \sys copies the victim page to the non-volatile
\texttt{\underline{shadow}} buffer (Line 5) and records the page's tag in a
non-volatile commit list (Lines 6-7). \sys uses the commit list to iterate over shadow pages 
at commit time, to copy the contents of shadow pages back to the main memory {\tt store}. 
%
%Dirty pages remain in {\tt \underline{shadow}} buffer until the
%dynamic task completes, at which point \sys commits the modified
%pages from {\tt \underline{shadow}} back to \texttt{\underline{store}}.
%
If the accessed page was previously modified and swapped out since the last
power failure (Line 8), the most recent version of the page is in {\tt
\underline{shadow}}, not \texttt{\underline{store}} (Line 9).

@@

@@

\textbf{Memory Consistency.} \sys's paging mechanism ensures that a task only
ever executes using consistent protected data. During task execution,
modifications to protected data do not affect their original memory locations,
because a task reads and writes the volatile \texttt{working} buffer only, and
modified pages are kept in \texttt{\underline{shadow}} until commit.  A power
failure erases the contents of the page buffer, preventing a re-executing task
from observing updates in the page buffer from a previous execution attempt.
Clearing \texttt{\underline{shadowCount}} as part of the second phase commit
(Line 14), ensures that all accesses to protected variables in subsequent tasks
correctly access their original memory locations in \texttt{\underline{store}}.
\sys assumes that the write to a non-volatile word in memory is atomic.
%
%The non-volatile commit bit persists
%across power failures and \sys repeatedly attempts to commit until it succeeds,
%despite arbitrary power failures.

%\begin{algorithm}[t]
% \begin{wrapfigure}{t}{0.5\textwidth}
% \fbox{\parbox{\linewidth}{
% 	\captionof{algorithm}{Two-phase commit}
% 	%\caption{Two-phase commit}
% 	\label{algo:commit}
% 	\scriptsize
\begin{algorithm}[t]
	\caption{Two-phase commit}
	\label{algo:commit}
	\scriptsize
	%\small
	\begin{algorithmic}[1]
        \Procedure{CommitPhase1}{} \Comment{On last coalesced task}
            \For {$i \in 0..|\texttt{working}|-1$}
                \State $t \gets \Call{Tag}{\texttt{working}[i]}$
                \State $\texttt{\underline{shadow}}[\Call{Tag}{\texttt{working}[i]}] \xleftarrow{\text{DMA}} \texttt{working}[i]$
                \State $\texttt{\underline{shadowList}}[\texttt{\underline{shadowCount}}] \gets t$
                \State $\texttt{\underline{shadowCount}} \gets \texttt{\underline{shadowCount}} + 1$
            \EndFor
            \State \Call{CommitPhase2}{}
        \EndProcedure
        \Procedure{CommitPhase2}{} \Comment{\texttt{\underline{commitIndex}} 0 on first boot}
            \State $\texttt{\underline{committing}} \gets \textbf{true}$
            \While{\texttt{\underline{commitIndex}} < \texttt{\underline{shadowCount}}}
                \State $t \gets \texttt{\underline{shadowList}}[\texttt{\underline{commitIndex}}]$
                \State \Call{commitToStore}{\texttt{\underline{shadow}}[t]}
                \State $\texttt{\underline{commitIndex}} \gets \texttt{\underline{commitIndex}} + 1$
            \EndWhile
            \State $\texttt{\underline{shadowCount}} \gets 0$
            \State $\texttt{\underline{commitIndex}} \gets 0$
            \State $\texttt{\underline{committing}} \gets \textbf{false}$
        \EndProcedure
        \Procedure{OnBoot}{} \Comment{Invoked on every boot}
            \If { \texttt{\underline{committing}} } \Call{CommitPhase2}{}
            \EndIf 
            \State $\texttt{\underline{shadowCount}} \gets 0$, $i_\text{victim} \gets 0$
        \EndProcedure
	\end{algorithmic}
\end{algorithm}
% 	\end{algorithmic}}}
% \end{wrapfigure}

\subsection{Dynamic Paging of \sys}
